Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-09 doc/invt.tex
intq.h: fixing bugs and improve stability

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { p_i }
  \equiv
  \alpha(t) \, \eta_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined a relative histogram
$\eta_i(t) \equiv h_i(t) /p_i$
in the last step.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, \eta_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

We can split the histogram into
a deterministic averaged part, $\langle \eta_i(t) \rangle$,
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  \eta_i(t) =
  \langle \eta_i(t) \rangle
  +
  \zeta_i(t)
  .
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle \eta_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { p_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { p_i }
  =
  \frac{                       e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_{*j}(t)} }
  \notag
  \\
  &\approx
  1 - v_{*i}(t) + \sum_{j = 1}^n p_j \, v_{*j}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n p_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n p_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average, we get
%
\begin{align}
  \langle \eta_{*i}(t) \rangle
  \equiv
  \langle \eta_i(t) \rangle -
  \sum_{j=1}^n \rho_j \langle \eta_j(t) \rangle
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its bin-average-shifted version, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac's delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.
%
For non-white noise,
the equivalent value of $G_i$ can be approximately obtained as
$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.
%
\note{More explicitly,
if the transition matrix of a period, $\tau$,
is $T^\tau_{ij}$,
then
$
  p_i p_j \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \, \pi_j - \pi_i \, \pi_j.
$
}%

From Eqs.
\eqref{eq:vt_diffeq}-\eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  =
  \alpha(t) \, \left[ \langle \eta_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is % of Eq. \eqref{eq:dxdt_singlebin} is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot y\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
y(q')
&\equiv
e^{q' - q(T)}.
\label{eq:y_def}
\end{align}

@@

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $v_{*i}(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $y(q(t))$
with fixed endpoints,
$y\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$y\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $y(q(t))$ should be a constant:
%
%We will further demand that
%%
%\begin{equation}
%  \lim_{t \to \infty} q(t) \to \infty.
%  \label{eq:qt_limit}
%\end{equation}
%%
%Then, for a large $t$,
%the second term on the right-hand side
%of Eq. \eqref{eq:xt_solution} can be neglected, and
%%
%\begin{align}
%\left\langle x^2(t) \right\rangle
%%&=
%%\int_0^t \int_0^t \dot y(t') \, \dot y(t'')
%%    \left\langle \xi(t') \xi(t'') \right\rangle dt'' \, dt'
%%\notag
%%\\
%&=
%\int_0^t \int_0^t
%  \dot y\bigl( q(t') \bigr) \,
%  \dot y\bigl( q(t'') \bigr) \,
%  \kappa(t' - t'') \, dt'' \, dt'.
%\label{eq:x2t_average_asym}
%\end{align}
%
%By the Euler-Lagrange equation, we get
%$$
%\begin{aligned}
%0
%&=
%\frac{d}{dt} \int_0^T
%  \dot y\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt'
%\\
%&= \int_0^T
%  \ddot y\bigl( q(t') \bigr) \, \kappa(t' - t) \, dt',
%\end{aligned}
%$$
%where we have dropped the boundary terms
%by assuming Eq. \eqref{eq:kappa_limit}
%and a large $T$.
%
%In fact, we can repeat the process $n$ times, and
%$$
%\int_0^t y^{(n)}(t') \, \kappa(t' - \tau) \, dt' = 0,
%\qquad (n \ge 1)
%$$
%
%In order for this to hold for any $t$,
%we must have
%%
%$
%\ddot y\bigl( q(t') \bigr) = 0
%,
%$
%or
\begin{equation}
  \dot y\bigl( q(t) \bigr) = c
  .
\label{eq:dydt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:y_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and the other constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

@@

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
The optimization procedure is similar to the single-bin case.
%
The key step comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.
%
%However, we will assume some restrictions and approximations
%to simplify the discussion.

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
and for the bin average of any quantity, $f$, we have
$\bav{f}= \mathcal F[f_i]_0 = \tilde f_0$,
so that
\begin{equation}
  \tilde f_{*k}
  = \tilde f_k - \bav{f} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde f_k  & k > 0,
  \end{dcases}
  \label{eq:ftstar}
\end{equation}
%
which means that the only the only effect
of bin average deduction is to
annihilate the first mode, $\tilde f_0$.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \eta}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \eta}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:xt2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot y_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  y_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \bigl\langle {\tilde \zeta}_{*k}(0)
\, {\tilde \zeta}_{*k}(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle {\tilde \zeta}_{*k}(0)
            \, {\tilde \zeta}_{*k}(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
We may understand Eq. \eqref{eq:Gamma_sum}
as an attempt of modeling
a general histogram fluctuation, ${\tilde \zeta}_{*k}(t)$,
as an equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 1}^{n-1}
%    \dot y_k\bigl( q(t) \bigr) \,
%    \dot y_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  y_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

%\subsubsection{White-noise approximation}

%In the asymptotic regime
%(i.e. if the simulation period, $T$, is sufficiently long),
%

%\subsection{\label{sec:mbin_opta}
%Optimal schedule for multiple-bin updating schemes}

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot y}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, y_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(Q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}\Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
\label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,
i.e., the asymptotic error grows
linearly with time at a rate of $\mathcal L$.
%
So we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
This serves
as a generalization of Eq. \eqref{eq:dydt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(Q)
    \;
    d Q
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    M( Q )
    \;
    d Q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.

@@

Based on the above results,
we now give a procedure of computing
the optimal schedule and the error
for a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_k$'s and $\Gamma_k$'s
from the given system and updating scheme (input),
and then apply them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
%The dependence of variables are shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process (MC or MD).
%
We have

>>>>

To proceed, we will
approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation\footnote{This
approximation requires us to shift the origin of time by $1$ in some cases.}
%
\begin{equation}
  \dot u_i(t)
  \equiv
  \frac{ d u_i(t) } { dt }
  \approx
  \alpha(t) \, \frac{ h_i(t) } { p_i }
  \equiv
  \alpha(t) \, \xi_i(t)
  ,
  \label{eq:ut_diffeq}
\end{equation}
%
with
%
$h_i(t) = \delta_{i, i(t)}$
%\begin{equation}
%  h_i(t) = \delta_{i, i(t)}
%  ,
%  \label{eq:h_def}
%\end{equation}
%
being the instantaneous histogram,
which is equal to $1$
for the bin $i(t)$ at time $t$
or zero otherwise,
and we have defined a relative histogram
$\xi_i(t) \equiv h_i(t) /p_i$
in the last step.
%
The shifted bias potential, $v_i(t)$,
follows the same differential equation
as the shift is a constant.
%
Deducting the bin average [cf. Eq. \eqref{eq:x_def}],
we get
%
\begin{equation}
  \dot v_{*i}(t)
  \approx
  \alpha(t) \, \xi_{*i}(t)
  .
  \label{eq:vt_diffeq}
\end{equation}

We can split the histogram into
a deterministic averaged part, $\langle \xi_i(t) \rangle$,
and a random fluctuation part, $\zeta_i(t)$,
%
\begin{equation}
  \xi_i(t) =
  \langle \xi_i(t) \rangle
  +
  \zeta_i(t)
  .
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to an average in an ensemble consists of
many copies of similar simulations
sharing the same schedule $\alpha(t)$
and the same bias potential at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.
%
For sufficiently small $\alpha(t)$,
the bias potential remains roughly the same for a short period,
and we may assume a quasi-equilibrium sampling process
such that
%with the deterministic part specified by Eq. \eqref{eq:pi_p_V}:
%
\begin{align}
  \langle \xi_i(t) \rangle
  %&=
  %\frac{ \langle h_i(t) } { p_i } \rangle
  &\approx
  \frac{ \pi_i(t) } { p_i }
  =
  \frac{                       e^{-v_{*i}(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_{*j}(t)} }
  \notag
  \\
  &\approx
  1 - v_{*i}(t) + \sum_{j = 1}^n p_j \, v_{*j}(t)
  ,
  \notag
  %\label{eq:h_ave}
\end{align}
%
where we have assumed the smallness
of $v_{*i}(t)$ in the linear expansion.
%
\note{
The second step follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - v_{*i}  }
     { \sum_{ r = 1 }^n p_j (1 - v_{*j}) }
=
\frac{                       1 - v_{*i}  }
     { 1 - \sum_{ j = 1 }^n p_j \, v_{*j} }
.
$$
}%
%
Deducting the bin average, we get
%
\begin{align}
  \langle \xi_{*i}(t) \rangle
  \equiv
  \langle \xi_i(t) \rangle -
  \sum_{j=1}^n \rho_j \langle \xi_j(t) \rangle
  &\approx
  - v_{*i}(t)
  .
  \label{eq:sh_ave}
\end{align}

We will approximate the histogram fluctuation part, $\zeta_i(t)$,
hence its bin-average-shifted version, $\zeta_{*i}(t)$,
as white noise such that
\begin{equation}
  \langle \zeta_{*i}(t) \, \zeta_{*i}(t') \rangle
  = G_i \, \delta(t - t')
  ,
  \label{eq:G_def}
\end{equation}
for some $G_i$,
and $\delta(t)$ is the Dirac's delta function.
%
We expect the approximation to hold for simulation
much longer than the autocorrelation time.
%
For non-white noise,
the equivalent value of $G_i$ can be approximately obtained as
$\sum_{t = -\infty}^\infty \langle \zeta_{*i}(t) \, \zeta_{*i}(0) \rangle$.

From Eqs.
\eqref{eq:vt_diffeq}-\eqref{eq:sh_ave},
we get % a set of decoupled equations
%
\begin{equation}
  \dot v_{*i}(t)
  =
  \alpha(t) \, \left[ \langle \xi_{*i}(t) \rangle + \zeta_{*i}(t) \right]
  \approx
  -\alpha(t) \, \left[ v_{*i}(t) - \zeta_{*i}(t) \right]
  ,
  \notag
  %\label{eq:dxdt_singlebin}
\end{equation}
%
and the formal solution is
%
\begin{equation}
  v_{*i}(T)
=
  v_{*i}(0) \, e^{-q(T)}
+
\int_0^T
  \dot y\bigl( q(t) \bigr) \, \zeta_{*i}(t) \, dt,
\notag
%\label{eq:xt_solution}
\end{equation}
%
where
%
$
q(t) \equiv \int_0^t \alpha(t') \, dt',
$
%
and
%
\begin{align}
y(q')
&\equiv
e^{q' - q(T)}.
\label{eq:y_def}
\end{align}

@@

The error depends implicitly on the curve, $\alpha(t)$,
or, equivalently, the curve, $q(t)$,
via $v_{*i}(T)$, and
our aim is to find the $q(t)$
that minimizes this error.
%
We now fix the endpoint value, $q(T)$,
and vary the curve, $q(t)$ ($0 < t < T$).
%
This is equivalent to varying the curve $y(q(t))$
with fixed endpoints,
$y\bigl( q(0) \bigr)  = e^{- q(T)}$
and
$y\bigl( q(T) \bigr) = 1$.
%
Note also that the first term on the right-hand side
of Eq. \eqref{eq:ET_average} is fixed during
the process,
while the second term is the action of a free particle.
Thus, the velocity of $y(q(t))$ should be a constant:
%
\begin{equation}
  \dot y\bigl( q(t) \bigr) = c
  .
\label{eq:dydt_const}
\end{equation}
%
Then, by using Eq. \eqref{eq:y_def}, we have
%
\begin{equation}
  e^{ q(t) - q(T) }
  =
  c \, (t + t_0)
  =
  \frac{ t + t_0 }
       { T + t_0 }
  ,
\label{eq:expqt}
\end{equation}
where $t_0$ is some constant,
and the other constant, $c$, has been determined
from the $t = T$ case
in the second step.
%
Taking the logarithm,
and differentiating both sides
with respect to $t$,
we get the inverse-time schedule
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }
  ,
\label{eq:alpha_invt1}
\end{equation}
%
and Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%

@@

We now turn to a general multiple-bin updating scheme.
%
By definition, Eq. \eqref{eq:mbin_update},
a visit to bin $j$ in this case results in updates of the bias potential
not only at bin $j$, but also at a few neighboring bins $i$'s.
%
The optimization procedure is similar to the single-bin case.
%
The key step comes from the
projection of the bias potential to the $n$ eigenmodes
of the updating matrix, $\mathbf w$,
formed by the relative magnitudes, $w_{ij}$'s.

@@

We can further define
a generalized Fourier transform, $\mathcal{F}$,
for variable, $A_i$,
%
\begin{equation}
  {\tilde A}_k
  \equiv \mathcal{F}[A_i]_k
  \equiv \sum_{i = 1}^n \phi_{ki} \, A_i
  ,
  \notag
  %\label{eq:gft_def}
\end{equation}
%
and the inverse transform is
%
\begin{equation}
  A_i =
  \frac{1}{\rho_i} \sum_{k = 1}^n \phi_{ki} \, \tilde{A}_k
  .
  \notag
  %\label{eq:gft_inv}
\end{equation}
%
Using this definition,
we can rewrite Eq. \eqref{eq:ortho1} as
$\mathcal F[1]_k = \delta_{k0}$,
and for the bin average of any quantity, $f$, we have
$\bav{f}= \mathcal F[f_i]_0 = \tilde f_0$,
so that
\begin{equation}
  \tilde f_{*k}
  = \tilde f_k - \bav{f} \, \delta_{k0}
  =
  \begin{dcases}
    0           & k = 0, \\
    \tilde f_k  & k > 0,
  \end{dcases}
  \label{eq:ftstar}
\end{equation}
%
which means that the only effect
of the bin-average deduction is to
annihilate the first mode, $\tilde f_0$.

We can now use Eq. \eqref{eq:eig_w} to diagonalize
the multiple-bin updating scheme, Eq. \eqref{eq:mbin_update},
as
%
$$
{\tilde v}_k(t + 1) =
{\tilde v}_k(t) + \alpha(t) \, \lambda_k \,
{\tilde \xi}_k(t)
,
$$
or in continuous time, we have, for $k > 0$,
%
\begin{equation}
  \dot{\tilde v}_k(t)
  =
  \alpha(t) \, \lambda_k \, {\tilde \xi}_k(t)
  \approx
  -\alpha(t) \, \lambda_k \,
  \bigl[ {\tilde v}_k(t) - {\tilde \zeta}_k(t) \bigr]
  ,
  \label{eq:ut_diffeq_mbin}
\end{equation}
%
where we have removed the bin average,
and used the transformed version of Eq. \eqref{eq:sh_ave}
[the subscript ``$_*$'' is dropped by Eq. \eqref{eq:ftstar}].
In the second step, we have used the linear approximation
as in Sec. \ref{sec:sbin_diffeq}.

@@

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:xt2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 1}^{n-1}
    \left\langle
      {\tilde v}_k^2(0)
    \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 1}^{n-1}
  \Gamma_k \, \dot y_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  y_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \bigl\langle {\tilde \zeta}_k(0)
\, {\tilde \zeta}_k(t) \bigr\rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle {\tilde \zeta}_k(0)
            \, {\tilde \zeta}_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
We may understand Eq. \eqref{eq:Gamma_sum}
as an attempt of modeling
a general histogram fluctuation, ${\tilde \zeta}_k(t)$,
as an equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we discuss several ways of estimating $\Gamma_k$'s.
%

@@

As in the single-bin case,
by varying the schedule, $\alpha(t)$, under a fixed value of
$q(T) = \int_0^T \alpha(t) \, dt$,
we can focus on the asymptotic error,
$\Err_A(T)$.
%
The value of $q(T)$ will be adjusted later
to minimize the total error in Sec. \ref{sec:optinitalpha}).
%
We can rewrite Eq. \eqref{eq:error_asym} much like an action
of a particle whose position is given by $q(t)$:
%
\begin{equation}
  \Err_A(T)
  =
  \int_0^T
    {\mathcal L} \bigl[ q(t)\bigr]
    \, dt
  ,
%\notag
\label{eq:error_asym_Lagrangian}
\end{equation}
%
where the Lagrangian is
%
\begin{align}
  {\mathcal L} \bigl[ q(t) \bigr]
  &=
  \sum_{ k = 1 }^{n - 1}
    \Gamma_k \, {\dot y}_k^2\bigl[ q(t) \bigr]
  %\notag
  %\\
  %&=
  =
  %\sum_{ k = 1 }^{n - 1}
  %  \Gamma_k \, \lambda_k^2 \, y_k^2 \bigl[ q(t) \bigr]
  %\; \dot q^2( t )
  %\notag
  %\\
  %&=
  M^2\bigl(q(T) - q(t) \bigr)
  \; \dot q^2( t )
  .
\notag
\end{align}
%
where we have
used Eq. \eqref{eq:uk_def} and
defined the (square-root) mass function as
%
\begin{equation}
  M(Q)
  \equiv
  \sqrt{
    \textstyle\sum_{ k = 1 }^{n - 1}\Gamma_k \, \lambda_k^2 \, e^{-2 \, \lambda_k \, Q}
  }
  .
  \notag
  %\label{eq:mass_func}
\end{equation}
%
In this mechanical analogy,
the schedule, $\dot q(t) = \alpha(t)$,
corresponds to the velocity of the particle,
%
and the above Lagrangian, containing only the kinetic energy,
describes a free particle
with a position-dependent mass.
%
Besides, the Hamiltonian
coincides with the Lagrangian:
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L
  .
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
Since the Lagrangian
does not explicitly depend on time, $t$,
the Hamiltonian is conserved,
which means that the asymptotic error grows
linearly with time at a rate of $\mathcal L$,
and we may set
%
\begin{equation}
  \sqrt{ \mathcal H }
  =
  \sqrt{ \mathcal L }
  =
  M\bigl( q(T) - q(t) \bigr)
  \;
  \dot q(t)
  =
  \frac{C}{T}
  ,
  \label{eq:Lagrangian_const}
\end{equation}
%
for some positive $C$.
%
This serves as a generalization of Eq. \eqref{eq:dydt_const}.
%
By integrating this equation of motion, we get
%
\begin{equation}
  \int_{ q(T) - q(t) }^{ q(T) }
    M(Q)
    \;
    d Q
  =
  C \, \frac t T
  ,
  \label{eq:q_opt}
\end{equation}
%
and $C$ can be determined from
the $t = T$ case as
%
\begin{equation}
  C =
  \int_{ 0 }^{ q(T) }
    M( Q )
    \;
    d Q
  .
  \label{eq:mint}
\end{equation}
%
Thus, we have an implicit equation for $q(t)$,
and hence the optimal schedule,
$\alpha(t) = \dot q(t)$.
%
The minimal asymptotic error is
%
\begin{equation}
  \Err_A(T)
  =
  \mathcal L \, T
  =
  \frac { C^2 } { T }
  =
  \frac 1 T
  \left(
    \int_0^{ q(T) } M(Q) \, dQ
  \right)^2
  .
\label{eq:error_asym2}
\end{equation}

@@

Based on the above results,
we now give a procedure of computing
the optimal schedule and the error
for a given adaptive FDS simulation.
%
Basically, we need to extract
$\lambda_k$'s and $\Gamma_k$'s
from the given system and updating scheme (input),
and then apply them to Eq. \eqref{eq:q_opt}
for the schedule,
as shown in Fig. \ref{fig:vardep}.
%
%The dependence of variables are shown in Fig. \ref{fig:vardep}.
%
While the updating scheme alone
determines $\lambda_k$'s,
$\Gamma_k$'s depend also on the system
and the underlying sampling process.
%
We have
